### 2021-06-14 papers/thesis/full/ch3-disk-slices/disk-slices.tex
(Hopefully almost) final edits made to thesis

For the disk, a review of spectral methods on the disk for solving the Poisson equation has been completed \cite{boyd2011comparing}, that looks at methods involving various bases for expanding functions in the disk including Zernike polynomials and Chebyshev polynomials. It was concluded that using Chebyshev polynomials yields mixed results, but that there was also no single advised basis to use, due to Zernike polynomials having a lack of radial fast transform at the time despite being found to yield accurate results. In the years since however, such radial fast transforms have been developed \cite{slevinsky2018use, sakai2009application, wilber2017computing}. The Zernike disk OPs can be written in terms of Jacobi polynomials \cite{boyd2011comparing}:
\bseqn{
	Z_n^m(r, \theta) := R_n^m(r) \sin(m\theta), \quad Z_n^{-m}(r, \theta) := R_n^m(r) \cos(m\theta),
}
where the radial part is
\bseqn{
	R_n^m(r) := r^m P^{(0,m)}_{(n-m)/2}(2r^2 - 1)
}
for $n - m$ even, while it is defined as $0$ for $n - m$ odd. A generalisation of the Zernike polynomials has been used for the development of a sparse spectral method on the disk \cite{vasil2016tensor}. These radial OPs are:
\bseqnnumber{
	Q_n^{k,m}(r) := C_n^{k,m} r^m P^{(k,m)}_n(2r^2 - 1), \label{eqn:ds:diskops}
}
where $C_n^{k,m}$ is some normalising constant so that they are orthonormal with respect to the inner product $\int_0^1 p(r) \; q(r) \; (1-r^2)^k \; r \; \D r$. $\{Q_n^{k,m}\}$ are therefore a two-parameter family of orthogonal polynomials on the unit disk. The classical Zernike polynomials are retrieved via setting $k = 0$. The use of polar coordinates here then allows for radially symmetric partial differential and other operators, as well as basis (parameter) conversion operators, to be represented as banded matrices by exploiting the relations of the Jacobi polynomials. Jacobi polynomials are extremely useful for solving a number of problems in both 1D and 2D, since one can utilise the sparsity of the relations that they hold to yield operator matrices that are efficient to solve \cite{doha2006efficient}.

>>>>

For the disk, a review of spectral methods on the disk for solving the Poisson equation has been completed \cite{boyd2011comparing}, that looks at methods involving various bases for expanding functions in the disk including Zernike polynomials and Chebyshev polynomials. It was concluded that using Chebyshev polynomials yields mixed results, but that there was also no single advised basis to use, due to Zernike polynomials having a lack of radial fast transform at the time despite being found to yield accurate results. In the years since however, such radial fast transforms have been developed \cite{slevinsky2018use, sakai2009application, wilber2017computing}. The Zernike disk OPs can be written in terms of Jacobi polynomials \cite{boyd2011comparing}:
\bseqn{
	Z_n^m(r, \theta) := R_n^m(r) \sin(m\theta), \quad Z_n^{-m}(r, \theta) := R_n^m(r) \cos(m\theta),
}
where the radial part is
\bseqn{
	R_n^m(r) := r^m P^{(0,m)}_{(n-m)/2}(2r^2 - 1)
}
for $n - m$ even, while it is defined as $0$ for $n - m$ odd. A generalisation of the Zernike polynomials has been used for the development of a sparse spectral method on the disk \cite{vasil2016tensor}. These radial OPs are:
\bseqnnumber{
	Q_n^{k,m}(r) := C_n^{k,m} r^m P^{(k,m)}_n(2r^2 - 1), \label{eqn:ds:diskops}
}
where $C_n^{k,m}$ is some normalising constant so that they are orthonormal with respect to the inner product $\int_0^1 p(r) \; q(r) \; (1-r^2)^k \; r \; \D r$. These OPs constitute a one-parameter family of orthogonal polynomials on the unit disk, with the classical Zernike polynomials being retrieved via setting the parameter $k = 0$. The use of polar coordinates allows for radially symmetric partial differential operators, while the incorporation of a parameter means they can be represented as banded matrices by exploiting the relations of the Jacobi polynomials. Jacobi polynomials are extremely useful for solving a number of problems in both 1D and 2D, since one can utilise the sparsity of the relations that they hold to yield operator matrices that are efficient to solve \cite{doha2006efficient}.

### 2021-06-14 papers/thesis/full/ch4-spherical-caps/spherical-caps.tex
Edits to Zernike stuff and other minor edits
(Hopefully almost) final edits made to thesis

	\medskip
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{sparsity-of-helmholtz}}
		\centering
		\caption{}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{sparsity-of-biharmonic}}
		\centering
		\caption{} 
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of (differential) operator matrices, showing their sparsity, plotted for $N = 20$. (a) The Laplace-Beltrami operator $\Delta^{(1)}_W$. (b) The $\rho^2$-factored Laplace-Beltrami operator $D_\varphi^{(0)} \: W_\varphi^{(1)} + T^{(0)\to(1)} \: T_W^{(1)\to(0)} \: (D_\theta)^2$. (c) The variable coefficient Helmholtz operator $\Delta^{(1)}_W + k^2 \: T^{(0)\to(1)} \: V({J_x^{(0)}}^\top, {J_y^{(0)}}^\top, {J_z^{(0)}}^\top) \: T_W^{(1)\to(0)}$ for $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$ and $k = 200$. (d) The Biharmonic operator $\calB_W^{(2)}$.}
	\label{fig:sc:sparsity}
\end{figure}

@@

Another desirable operator is the Biharmonic operator $\DeltaS^2$, for which we assume zero Dirichlet and Neumann conditions. That is, \bstodo{All functions not dependent on r (ie any we expand in this basis) satisfy the derivative BC automatically. I am fairly sure this is still the BC for the sphere though..?}
\bseqn{
	u(x,y,z) &= 0, \quad \pfpx{n}{u}(x,y,z) = \nabla_S u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 \quad \text{for } (x,y,z) \in \partial\Omega
}
where $\partial\Omega$ is the $z=\alpha$ boundary, and $\unitvec{n}(x,y,z)$ is the outward unit normal vector at the point $(x,y,z)$ on the boundary, i.e. $\unitvec{n}(x,y,z) = \unitvec{n}(\vec x) := {\vec x \over \norm{\vec x}} = \vec x$. The matrix operator for the Biharmonic operator will take us from coefficients in the space $\bigW^{(2)}(x,y,z)$ to coefficients in the space $\bigscopN^{(2)}(x,y,z)$. To construct this, we can simply multiply together two of the spherical Laplacian operators defined in \bsrefdef{def:sc:differentialoperators}, namely $\calL^{(0)\to(2)}$ and $\calL_W^{(2)\to(0)}$:
\bseqn{
	\calB^{(2)}_W := \calL^{(0)\to(2)} \: \calL_W^{(2)\to(0)}.
}
Since the operator $\calL_W^{(2)\to(0)}$ acts on coefficients in the $\bigW^{(2)}(x,y,z)$ space, we ensure that we satisfy the zero Dirichlet and Neumann boundary conditions -- such a function could be written $u(x,y,z) = \genjacw^{(2,0)}(z) \: \tilde u(x,y,z)$ and thus its spherical gradient would be zero on the boundary $z = \alpha$. This allows us to apply the $\calL^{(0)\to(2)}$ operator after, safe in the knowledge that boundary conditions have been accounted for. The sparsity and structure of this biharmonic operator are seen in \bsreffig{fig:sc:sparsity}.

@@

We now demonstrate how the sparse linear systems constructed as above can be used to efficiently solve PDEs with zero Dirichlet conditions on the spherical cap defined by $\Omega$. We consider Poisson, inhomogeneous variable coefficient Helmholtz equation and the Biharmonic equation, as well as a time dependent heat equation, demonstrating the versatility of the approach.

@@

Our last example is the biharmonic equation: find $u(x,y,z)$ given a function $f(x,y,z)$ such that:
\bseqnnumber{
	\begin{cases}
    		\DeltaS^2 u(x,y,z) = f(x,y,z) &\quad \text{in } \Omega \\
		u(x,y,z) = 0, \quad \frac{\partial u}{\partial n}(x,y,z) = \grad_S \: u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 &\quad \text{on } \partial \Omega
	\end{cases}
	\label{eqn:sc:biharmonic}
}
where $\DeltaS^2$ is the Biharmonic operator, noting the imposition of zero Dirichlet and Neumann boundary conditions on $u$. For clarity, we reiterate that the unit normal vector in this sense is simply $\unitvec{n}(x,y,z) = \unitvec{n}(\vec x) := {\vec x \over \norm{\vec x}} = \vec x$ (see \bsrefsection{section:sc:furtherdiffoperators}). In \bsreffig{fig:sc:biharmonic} we see the solution to the Biharmonic \bsrefeqn{eqn:sc:biharmonic} in the spherical cap $\Omega$. In \bsreffig{fig:sc:biharmonic} we also show the norms of each block of calculated coefficients of the approximation for four more complex right-hand sides of the biharmonic equation with $N = 200$, that is, $(N+1)^2 = 40,401$ unknowns. Once again, the rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution, and we see that we achieve exponential convergence for these more complex functions.

>>>>

	\medskip
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{sparsity-of-helmholtz}}
		\centering
		\caption{}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{sparsity-of-biharmonic}}
		\centering
		\caption{} 
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of (differential) operator matrices, showing their sparsity, plotted for $N = 20$. Note that here each block corresponds to the order $k$ and not the degree $n$. (a) The Laplace-Beltrami operator $\Delta^{(1)}_W$. (b) The $\rho^2$-factored Laplace-Beltrami operator $D_\varphi^{(0)} \: W_\varphi^{(1)} + T^{(0)\to(1)} \: T_W^{(1)\to(0)} \: (D_\theta)^2$. (c) The variable coefficient Helmholtz operator $\Delta^{(1)}_W + k^2 \: T^{(0)\to(1)} \: V({J_x^{(0)}}^\top, {J_y^{(0)}}^\top, {J_z^{(0)}}^\top) \: T_W^{(1)\to(0)}$ for $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$ and $k = 200$. (d) The biharmonic operator $\calB_W^{(2)}$.}
	\label{fig:sc:sparsity}
\end{figure}

@@

Another desirable operator is the biharmonic operator $\DeltaS^2$, for which we assume zero Dirichlet and Neumann conditions. That is, 
\bseqn{
	u(x,y,z) &= 0, \quad \pfpx{n}{u}(x,y,z) = \gradS u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 \quad \text{for } (x,y,z) \in \partial\Omega
}
where $\partial\Omega$ is the $z=\alpha$ boundary, and $\unitvec{n}(x,y,z)$ is the outward unit normal vector at the point $(x,y,z)$ on the boundary. For the spherical cap $\Omega$, this is simply $\phivec(x,y,z)$. Hence, 
\bseqn{
	\gradS u(x,y,z) \cdot \unitvec{n}(x,y,z) = \gradS u(x,y,z) \cdot \phivec(x,y,z) = - \rho(z) \pfpx{z}{u}(x,y,z)
} 
and the boundary conditions simplify to
\bseqn{
	u(x,y,z) &= 0, \quad \pfpx{z}{u}(x,y,z) = 0 \quad \text{for } (x,y,z) \in \partial\Omega.
}
The matrix operator for the biharmonic operator will take us from coefficients in the space $\bigW^{(2)}(x,y,z)$ to coefficients in the space $\bigscopN^{(2)}(x,y,z)$. To construct this, we can simply multiply together two of the spherical Laplacian operators defined in \bsrefdef{def:sc:differentialoperators}, namely $\calL^{(0)\to(2)}$ and $\calL_W^{(2)\to(0)}$:
\bseqn{
	\calB^{(2)}_W := \calL^{(0)\to(2)} \: \calL_W^{(2)\to(0)}.
}
Since the operator $\calL_W^{(2)\to(0)}$ acts on coefficients in the $\bigW^{(2)}(x,y,z)$ space, we ensure that we satisfy the zero Dirichlet and Neumann boundary conditions -- such a function could be written $u(x,y,z) = \genjacw^{(2,0)}(z) \: \tilde u(x,y,z)$. This allows us to apply the $\calL^{(0)\to(2)}$ operator after, safe in the knowledge that boundary conditions have been accounted for. The sparsity and structure of this biharmonic operator are seen in \bsreffig{fig:sc:sparsity}.

@@

We now demonstrate how the sparse linear systems constructed as above can be used to efficiently solve PDEs with zero Dirichlet conditions on the spherical cap defined by $\Omega$. We consider Poisson, inhomogeneous variable coefficient Helmholtz equation and the biharmonic equation, as well as a time dependent heat equation, demonstrating the versatility of the approach.

@@

Our last example is the biharmonic equation. Find $u(x,y,z)$ given a function $f(x,y,z)$ such that:
\bseqnnumber{
	\begin{cases}
    		\DeltaS^2 u(x,y,z) = f(x,y,z) &\quad \text{in } \Omega \\
		u(x,y,z) = 0, \quad \frac{\partial u}{\partial n}(x,y,z) = \grad_S \: u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 &\quad \text{on } \partial \Omega
	\end{cases}
	\label{eqn:sc:biharmonic}
}
where $\DeltaS^2$ is the biharmonic operator, noting the imposition of zero Dirichlet and Neumann boundary conditions on $u$. Recall from \bsrefsection{section:sc:furtherdiffoperators} that for the spherical cap, the boundary conditions here simplify to $u = \pfpx{z}{u} = 0$ on $\partial \Omega$ (since the normal vector of the boundary is simply $\phivec(x,y,z)$). In \bsreffig{fig:sc:biharmonic} we see the solution to the biharmonic \bsrefeqn{eqn:sc:biharmonic} in the spherical cap $\Omega$. In \bsreffig{fig:sc:biharmonic} we also show the norms of each block of calculated coefficients of the approximation for four more complex right-hand sides of the biharmonic equation with $N = 200$, that is, $(N+1)^2 = 40,401$ unknowns. Once again, the rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution, and we see that we achieve exponential convergence for these more complex functions.

### 2021-06-14 papers/thesis/full/conclusion/summary.tex
(Hopefully almost) final edits made to thesis

Discretizations of partial differential operators are sparse on other suitable sub-domains of the sphere, as evidenced by the construction presented for the spherical cap. Hence, using the techniques learnt here, one could also look to constructing similar methods for solving PDEs on spherical triangles, with a view to using these sub-domains as elements for a spherical spectral element method. Moreover, one could use this framework to also develop sparse spectral methods on more general domains defined by an algebraic surface -- in particular those defined by non-uniform rational B-splines (NURBS) \bstodo{Not sure what to say exactly or how in depth to go here}. However, it is not obvious how one would derive or compute the OPs for such domains.

>>>>

Discretizations of partial differential operators are sparse on other suitable sub-domains of the sphere, as evidenced by the construction presented for the spherical cap. Hence, using the techniques learnt here, one could also look to constructing similar methods for solving PDEs on spherical triangles, with a view to using these sub-domains as elements for a spherical spectral element method. Moreover, one could use this framework to also develop sparse spectral methods on more general domains defined by an algebraic surface -- in particular those defined by non-uniform rational B-splines (NURBS). However, it is not obvious how one would derive or compute the OPs for such domains.
